pipeComp, a general framework for the evaluation of computational pipelines, reveals performant single cell RNA-seq preprocessing tools

### 2020-03-02 bmc_article.tex
Updates from Overleaf

\iffalse

\begin{figure}
    \centering
    \includegraphics[width=\textwidth,keepaspectratio]{dataset_description}
    \caption{\textbf{Overview of the benchmark datasets used.}}
    \label{fig:figure1}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[scale=0.7, keepaspectratio]{main_figures/pipeline_explanation.pdf}
    \caption{\textbf{Overview of the \textit{pipeComp} framework and its application to a scRNAseq clustering pipeline. A:} The package is built around a \texttt{PipelineDefinition} S4 class which defines a set of functions to be executed consecutively, as well as optional evaluation functions for each step. Each step can accept a number of parameters whose alternative values are provided as a list. All (or subsets of) combinations of parameters can then be simultaneously ran and evaluated using the \texttt{runPipeline} function. \textbf{B:} Scheme representing the application of \textit{pipeComp} to evaluate a range of methods that are commonly used in scRNAseq studies and some of the metrics monitored at various steps.}
    \label{fig:figure2}
\end{figure}

\fi

@@

\iffalse
\begin{figure}
    \centering
    \includegraphics[scale=0.65, keepaspectratio]{{main_figures/doublets.pdf}}
    \caption{\textbf{Identification of doublet cells. A:} Receiver operating characteristic (ROC) curves of the tested doublet detection methods for three datasets with SNP-identified doublets. Dots indicate threshold determined by the true number of doublets. \textbf{B:} Running time of the different methods (\textit{DoubletFinder} failed on one of the datasets). \textbf{C:} Rate of misclassification of the cells identified by \textit{scDblFinder} as doublets (DBL) or singlets (SNG), across a large range of clustering analysis. Even in datasets which should not have neotypic doublets, the cells identified as such tend to be misclassified.    \label{fig:figure3}}
\end{figure}
\fi

@@

\iffalse
\begin{figure}
    \centering
    \includegraphics[scale=0.60, keepaspectratio]{{main_figures/filtering.pdf}}
    \caption{\textbf{Effect of filtering on cell subpopulation structure and clustering. A:} Filtering on the basis of distance to the whole distribution can lead to strong bias against certain subpopulations. The dashed line indicates a threshold of 2.5 median absolute deviations (MADs) from the median of the overall population. \textbf{B:} Relationship between the maximum subpopulation exclusion rate and the average clustering accuracy per subpopulation across various filtering strategies. Of note, doublet removal appears to be desirable even when, due to the design, there are no heterotypic doublets in the data. The PCA methods refer to multivariate outlier detected as implemented in \textit{scater} (see Methods for details).}
    \label{fig:figure4}
\end{figure}
\fi

@@

\iffalse
\begin{figure}
    \centering
    \includegraphics[scale=0.55, keepaspectratio]{{main_figures/normalization.pdf}}
    \caption{\textbf{Evaluation of normalization procedures.} \textbf{A:} Variance in PC1 explained by library size (left) and detection rate (right) after accounting for subpopulation differences. \textbf{B:} Average silhouette width per true subpopulation, where higher silhouette width means a higher separability. \textbf{C:} Clustering accuracy (\textit{Seurat} clustering), measured by mutual information (MI), Adjusted Rand Index (ARI) and ARI at the true number of cluster. 'feat_regress', 'mt_regress' and 'feat_mt_regress' respectively stand for the regressing out of the number of features, the proportion of mitochondrial reads, or both during scaling. }
    \label{fig:figure5}
\end{figure}
\fi

@@

\iffalse
\begin{figure}
    \centering
    \includegraphics[scale=0.65]{{main_figures/selection.pdf}}
    \caption{\textbf{Evaluation of feature selection methods. A:} Ability of different feature ranking methods to capture genes with a high proportion of variance (left) or deviance (right) explained by real subpopulations. The asterisk denotes the default Seurat method. \textbf{B:} Accuracy of clusterings (at the true number of clusters) when selecting 1000 genes using the given methods. Based on standard \textit{Seurat} normalization (left) or \textit{sctransform} (right). The  \textit{vst.varExp} and \textit{devianceExplained} methods correspond to the estimates used in \textbf{A} to evaluate the selection methods and were included here only for validation purpose.}
    \label{fig:figure6}
\end{figure}
\fi

@@

\iffalse
\begin{figure}
    \centering
    \includegraphics[scale=0.40, keepaspectratio]{{main_figures/dr.pdf}}
    \caption{\textbf{Evaluation of common dimensionality reduction methods. A:} Average silhouette width per subpopulation resulting from combinations of normalization and dimension reductions. \textbf{B:} Clustering accuracy, mean proportion of the variance in the first 5 components explained by real subpopulations (center), and median ARI of the resulting clustering (right).}
    \label{fig:figure7}
\end{figure}
\fi

@@

\iffalse
\begin{figure}
    \centering
    \includegraphics[scale=0.53, keepaspectratio]{{main_figures/dimensionality.pdf}}
    \caption{\textbf{Estimating dimensionality. A:} Estimated dimensionality using the proportion of variance in each component explained by the true subpopulations. \textbf{B:} Difference between `real' dimensionality (from \textbf{A}) and dimensionality estimation methods, along with computing time. \textbf{C:} Average silhouette width per subpopulation across a selection of methods, combined with \textit{sctransform} or standard \textit{Seurat} normalization. \textbf{D:} Clustering accuracy across the normalization/dimensionality estimation methods.}
    \label{fig:figure8}
\end{figure}
\fi

@@

\iffalse
\begin{figure}
    \centering
    \includegraphics[scale=0.34, keepaspectratio]{{main_figures/clustering.pdf}}
    \caption{\textbf{Evaluation of clustering methods.} Clustering accuracy (\textbf{A}) and run time (\textbf{B}) of \textit{scran}- and \textit{Seurat}-based clustering tools in combination with \textit{sctransform} and standard \textit{Seurat}'s normalization.}
    \label{fig:figure9}
\end{figure}
\fi

@@

\iffalse
\begin{figure}
    \centering
    \includegraphics[scale=0.49, keepaspectratio]{{main_figures/imputation.pdf}}
     \caption{\textbf{Evaluation of imputation/denoising methods} Average silhouette width per subpopulation (\textbf{A}) and clustering accuracy (\textbf{B}) with or without (indicated as \textit{none}) application of a denoising/imputation method.}
    \label{fig:figure10}
\end{figure}
\fi

@@

\iffalse
\begin{figure}[h]
    \centering
    \includegraphics[scale=0.9,keepaspectratio]{main_figures/pipeComp_summary.pdf}
    \caption{\textbf{Recommendations of tools from filtering to clustering.}}
    \label{fig:figure11}
\end{figure}
\fi

@@

Our recommendations are in line with the results of several other benchmarking studies. For instance, \textit{Seurat} and \textit{scran} methods were already shown to yield good clustering performance \cite{freytagComparison2018} and \textit{DrImpute} to improve the quality of downstream analyses \cite{ZhangImput2018}. A previous systematic comparison \cite{viethSystematic2019} instead found that the use of \textit{SAVER} for denoising data and \textit{scran} for clustering yielded better performance on the ability to recover differentially-expressed genes, whereas we focused on the ability to recover cell populations. Another disagreement with the present recommendations come from the study of Hou et al. \cite{HouImput2020} that recommends \textit{SAVERX} and discourage the use of \textit{DCA} for better clustering, whereas our study found opposite results for these tools. Several differences in the design of our study could explain this contrast, such as the use of different metrics (ARI at true number of clusters, mean precision/ recall), of different datasets (Koh, Kumar, simMix simulations) and the use of different filtering and normalization strategies. \textit{ALRA} however yielded good results in both studies and may be the most conservative option. The study from Hou et al. also highlighted the usefulness of \textit{ALRA} when followed by differential and trajectory analysis both in UMI and Fluidigm technologies.

@@

We developed \texttt{pipeComp} to address the need of a framework that can simultaneously evaluate the interaction of multiple tools. The work of Vieth and colleagues\cite{viethSystematic2019} already offered an important precedent in this respect, evaluating the interaction of various steps in the context of scRNAseq differential expression analysis, but did not offer a platform for doing so. Instead, the \textit{CellBench} package was recently proposed to address a similar need \cite{suCellbench}, offering an elegant piping syntax to combine alternative methods at successive steps. A key additional feature of \textit{pipeComp} is its ability to perform evaluation in parallel to the pipeline and offering on-the-fly evaluation at any step. This avoids the need to store potentially large intermediate data for all possible permutations and thus allowing a combinatorial benchmark. In addition, the fact that benchmark functions are stored in the \texttt{PipelineDefinition} makes the benchmark more smoothly portable, making it easy to modify, extend with further methods, or apply to other datasets.

@@

All analyses were performed through the \href{https://github.com/plger/pipeComp}{pipeComp} R package, which implements the pipeline framework described here. scVI results and results of the clustering section were produced with pipeComp 0.99.4, while all other results were produced with pipeComp 0.99.2 (\textit{biorxiv}) branch of the repository). All code to reproduce the simulations and figures is available in the \url{https://github.com/markrobinsonuzh/scRNA\_pipelines\_paper} repository, which also includes the basic datasets with a standardized annotation.

@@

The deviance explained by subpopulations was calculated directly on counts using the \linebreak \texttt{getDevianceExplained} function of \texttt{pipeComp}. The function uses \textit{edgeR} to fit two models, a full ($\sim librarySize + population$) and a reduced one ($\sim librarySize$). For each gene, the deviance explained is then the difference between the deviance of each models, divided by the deviance of the reduced model. In the rare cases where this resulted in a negative deviance explained, it was set to zero.

@@

\begin{figure}[h!]
    \centering
    \caption{\textbf{Overview of the benchmark datasets used.}}
    \label{fig:figure1}
\end{figure}

\begin{figure}[h!]
    \centering
    \caption{\textbf{Overview of the \textit{pipeComp} framework and its application to a scRNAseq clustering pipeline. A:} The package is built around a \texttt{PipelineDefinition} S4 class which defines a set of functions to be executed consecutively, as well as optional evaluation functions for each step. Each step can accept a number of parameters whose alternative values are provided as a list. All (or subsets of) combinations of parameters can then be simultaneously ran and evaluated using the \texttt{runPipeline} function. \textbf{B:} Scheme representing the application of \textit{pipeComp} to evaluate a range of methods that are commonly used in scRNAseq studies and some of the metrics monitored at various steps.}
    \label{fig:figure2}
\end{figure}

\begin{figure}[h!]
    \centering
    \caption{\textbf{Identification of doublet cells. A:} Receiver operating characteristic (ROC) curves of the tested doublet detection methods for three datasets with SNP-identified doublets. Dots indicate threshold determined by the true number of doublets. \textbf{B:} Running time of the different methods (\textit{DoubletFinder} failed on one of the datasets). \textbf{C:} Rate of misclassification of the cells identified by \textit{scDblFinder} as doublets (DBL) or singlets (SNG), across a large range of clustering analysis. Even in datasets which should not have neotypic doublets, the cells identified as such tend to be misclassified.    \label{fig:figure3}}
\end{figure}

\begin{figure}[h!]
    \centering
    \caption{\textbf{Effect of filtering on cell subpopulation structure and clustering. A:} Filtering on the basis of distance to the whole distribution can lead to strong bias against certain subpopulations. The dashed line indicates a threshold of 2.5 median absolute deviations (MADs) from the median of the overall population. \textbf{B:} Relationship between the maximum subpopulation exclusion rate and the average clustering accuracy per subpopulation across various filtering strategies. Of note, doublet removal appears to be desirable even when, due to the design, there are no heterotypic doublets in the data. The PCA methods refer to multivariate outlier detected as implemented in \textit{scater} (see Methods for details).}
    \label{fig:figure4}
\end{figure}

\begin{figure}[h!]
    \centering
    \caption{\textbf{Evaluation of normalization procedures.} \textbf{A:} Variance in PC1 explained by library size (left) and detection rate (right) after accounting for subpopulation differences. \textbf{B:} Average silhouette width per true subpopulation, where higher silhouette width means a higher separability. \textbf{C:} Clustering accuracy (\textit{Seurat} clustering), measured by mutual information (MI), Adjusted Rand Index (ARI) and ARI at the true number of cluster. 'feat_regress', 'mt_regress' and 'feat_mt_regress' respectively stand for the regressing out of the number of features, the proportion of mitochondrial reads, or both during scaling. }
    \label{fig:figure5}
\end{figure}

\begin{figure}[h!]
    \centering
    \caption{\textbf{Evaluation of feature selection methods. A:} Ability of different feature ranking methods to capture genes with a high proportion of variance (left) or deviance (right) explained by real subpopulations. The asterisk denotes the default Seurat method. \textbf{B:} Accuracy of clusterings (at the true number of clusters) when selecting 1000 genes using the given methods. Based on standard \textit{Seurat} normalization (left) or \textit{sctransform} (right). The  \textit{vst.varExp} and \textit{devianceExplained} methods correspond to the estimates used in \textbf{A} to evaluate the selection methods and were included here only for validation purpose.}
    \label{fig:figure6}
\end{figure}

\begin{figure}[h!]
    \centering
    \caption{\textbf{Evaluation of common dimensionality reduction methods. A:} Average silhouette width per subpopulation resulting from combinations of normalization and dimension reductions. \textbf{B:} Clustering accuracy, mean proportion of the variance in the first 5 components explained by real subpopulations (center), and median ARI of the resulting clustering (right).}
    \label{fig:figure7}
\end{figure}

\begin{figure}[h!]
    \centering
    \caption{\textbf{Estimating dimensionality. A:} Estimated dimensionality using the proportion of variance in each component explained by the true subpopulations. \textbf{B:} Difference between `real' dimensionality (from \textbf{A}) and dimensionality estimation methods, along with computing time. \textbf{C:} Average silhouette width per subpopulation across a selection of methods, combined with \textit{sctransform} or standard \textit{Seurat} normalization. \textbf{D:} Clustering accuracy across the normalization/dimensionality estimation methods.}
    \label{fig:figure8}
\end{figure}

\begin{figure}[h!]
    \centering
    \caption{\textbf{Evaluation of clustering methods.} Clustering accuracy (\textbf{A}) and run time (\textbf{B}) of \textit{scran}- and \textit{Seurat}-based clustering tools in combination with \textit{sctransform} and standard \textit{Seurat}'s normalization.}
    \label{fig:figure9}
\end{figure}

\begin{figure}[h!]
    \centering
    \caption{\textbf{Evaluation of imputation/denoising methods} Average silhouette width per subpopulation (\textbf{A}) and clustering accuracy (\textbf{B}) with or without (indicated as \textit{none}) application of a denoising/imputation method.}
    \label{fig:figure10}
\end{figure}

\begin{figure}[h!]
    \centering
    \caption{\textbf{Recommendations of tools from filtering to clustering.}}
    \label{fig:figure11}
\end{figure}

>>>>

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

Our recommendations are in line with the results of several other benchmarking studies. For instance, \textit{Seurat} and \textit{scran} methods were already shown to yield good clustering performance \cite{freytagComparison2018} and \textit{DrImpute} to improve the quality of downstream analyses \cite{ZhangImput2018}. A previous systematic comparison \cite{viethSystematic2019} instead found that the use of \textit{SAVERX} for denoising data and \textit{scran} for clustering yielded better performance on the ability to recover differentially-expressed genes, whereas we focused on the ability to recover cell populations. Another disagreement with the present recommendations come from the study of Hou et al. \cite{HouImput2020} that recommends \textit{SAVERX} and discourage the use of \textit{DCA} for better clustering, whereas our study found opposite results for these tools. Several differences in the design of our study could explain this contrast, such as the use of different metrics (ARI at true number of clusters, mean precision/ recall), of different datasets (Koh, Kumar, simMix simulations) and the use of different filtering and normalization strategies. \textit{ALRA} however yielded good results in both studies and may be the most conservative option. The study from Hou et al. also highlighted the usefulness of \textit{ALRA} when followed by differential and trajectory analysis both in UMI and Fluidigm technologies.

@@

We developed \textit{pipeComp} to address the need of a framework that can simultaneously evaluate the interaction of multiple tools. The work of Vieth and colleagues\cite{viethSystematic2019} already offered an important precedent in this respect, evaluating the interaction of various steps in the context of scRNAseq differential expression analysis, but did not offer a platform for doing so. Instead, the \textit{CellBench} package was recently proposed to address a similar need \cite{suCellbench}, offering an elegant piping syntax to combine alternative methods at successive steps. A key additional feature of \textit{pipeComp} is its ability to perform evaluation in parallel to the pipeline and offering on-the-fly evaluation at any step. This avoids the need to store potentially large intermediate data for all possible permutations and thus allowing a combinatorial benchmark. In addition, the fact that benchmark functions are stored in the \texttt{PipelineDefinition} makes the benchmark more smoothly portable, making it easy to modify, extend with further methods, or apply to other datasets.

@@

All analyses were performed through the \href{https://github.com/plger/pipeComp}{pipeComp} R package, which implements the pipeline framework described here. scVI results and results of the clustering section were produced with \textit{pipeComp} 0.99.4, while all other results were produced with version 0.99.2 (\textit{biorxiv}) branch of the repository). All code to reproduce the simulations and figures is available in the \url{https://github.com/markrobinsonuzh/scRNA\_pipelines\_paper} repository, which also includes the basic datasets with a standardized annotation.

@@

The deviance explained by subpopulations was calculated directly on counts using the \linebreak \texttt{getDevianceExplained} function of \textit{pipeComp}. The function uses \textit{edgeR} to fit two models, a full ($\sim librarySize + population$) and a reduced one ($\sim librarySize$). For each gene, the deviance explained is then the difference between the deviance of each models, divided by the deviance of the reduced model. In the rare cases where this resulted in a negative deviance explained, it was set to zero.

@@

\showfigstrue

\begin{figure}
    \centering
    \ifshowfigs
    \includegraphics[width=\textwidth,keepaspectratio]{dataset_description}
    \fi
    \caption{\textbf{Overview of the benchmark datasets used.}}
    \label{fig:figure1}
\end{figure}

\begin{figure}
    \centering
    \ifshowfigs
    \includegraphics[scale=0.75, keepaspectratio]{main_figures/pipeline_explanation.pdf}
    \fi
    \caption{\textbf{Overview of the \textit{pipeComp} framework and its application to a scRNAseq clustering pipeline. A:} The package is built around a \texttt{PipelineDefinition} S4 class which defines a set of functions to be executed consecutively, as well as optional evaluation functions for each step. Each step can accept a number of parameters whose alternative values are provided as a list. All (or subsets of) combinations of parameters can then be simultaneously ran and evaluated using the \texttt{runPipeline} function. \textbf{B:} Scheme representing the application of \textit{pipeComp} to evaluate a range of methods that are commonly used in scRNAseq studies and some of the metrics monitored at various steps.}
    \label{fig:figure2}
\end{figure}

\begin{figure}
    \centering
    \ifshowfigs
    \includegraphics[scale=0.85, keepaspectratio]{{main_figures/doublets.pdf}}
    \fi
    \caption{\textbf{Identification of doublet cells. A:} Receiver operating characteristic (ROC) curves of the tested doublet detection methods for three datasets with SNP-identified doublets. Dots indicate threshold determined by the true number of doublets. \textbf{B:} Running time of the different methods (\textit{DoubletFinder} failed on one of the datasets). \textbf{C:} Rate of misclassification of the cells identified by \textit{scDblFinder} as doublets (DBL) or singlets (SNG), across a large range of clustering analysis. Even in datasets which should not have neotypic doublets, the cells identified as such tend to be misclassified.    \label{fig:figure3}}
\end{figure}

\begin{figure}
    \centering
    \ifshowfigs
    \includegraphics[scale=0.72, keepaspectratio]{{main_figures/filtering.pdf}}
    \fi
    \caption{\textbf{Effect of filtering on cell subpopulation structure and clustering. A:} Filtering on the basis of distance to the whole distribution can lead to strong bias against certain subpopulations. The dashed line indicates a threshold of 2.5 median absolute deviations (MADs) from the median of the overall population. \textbf{B:} Relationship between the maximum subpopulation exclusion rate and the average clustering accuracy per subpopulation across various filtering strategies. Of note, doublet removal appears to be desirable even when, due to the design, there are no heterotypic doublets in the data. The PCA methods refer to multivariate outlier detected as implemented in \textit{scater} (see Methods for details).}
    \label{fig:figure4}
\end{figure}

\begin{figure}
    \centering
    \ifshowfigs
    \includegraphics[scale=0.6, keepaspectratio]{{main_figures/normalization.pdf}}
    \fi
    \caption{\textbf{Evaluation of normalization procedures. A:} Variance in PC1 explained by library size (left) and detection rate (right) after accounting for subpopulation differences. \textbf{B:} Average silhouette width per true subpopulation, where higher silhouette width means a higher separability. \textbf{C:} Clustering accuracy (\textit{Seurat} clustering), measured by mutual information (MI), Adjusted Rand Index (ARI) and ARI at the true number of cluster. `feat\_regress', `mt\_regress' and `feat\_mt\_regress' respectively stand for the regressing out of the number of features, the proportion of mitochondrial reads, or both during scaling.}
    \label{fig:figure5}
\end{figure}

\begin{figure}
    \centering
    \ifshowfigs
    \includegraphics[scale=0.65]{{main_figures/selection.pdf}}
    \fi
    \caption{\textbf{Evaluation of feature selection methods. A:} Ability of different feature ranking methods to capture genes with a high proportion of variance (left) or deviance (right) explained by real subpopulations. The asterisk denotes the default Seurat method. \textbf{B:} Accuracy of clusterings (at the true number of clusters) when selecting 1000 genes using the given methods. Based on standard \textit{Seurat} normalization (left) or \textit{sctransform} (right). The  \textit{vst.varExp} and \textit{devianceExplained} methods correspond to the estimates used in \textbf{A} to evaluate the selection methods and were included here only for validation purpose.}
    \label{fig:figure6}
\end{figure}

\begin{figure}
    \centering
    \ifshowfigs
    \includegraphics[scale=0.50, keepaspectratio]{{main_figures/dr.pdf}}
    \fi
    \caption{\textbf{Evaluation of common dimensionality reduction methods. A:} Average silhouette width per subpopulation resulting from combinations of normalization and dimension reductions. \textbf{B:} Clustering accuracy, mean proportion of the variance in the first 5 components explained by real subpopulations (center), and median ARI of the resulting clustering (right).}
    \label{fig:figure7}
\end{figure}

\begin{figure}
    \centering
    \ifshowfigs
    \includegraphics[scale=0.53, keepaspectratio]{{main_figures/dimensionality.pdf}}
    \fi
    \caption{\textbf{Estimating dimensionality. A:} Estimated dimensionality using the proportion of variance in each component explained by the true subpopulations. \textbf{B:} Difference between `real' dimensionality (from \textbf{A}) and dimensionality estimation methods, along with computing time. \textbf{C:} Average silhouette width per subpopulation across a selection of methods, combined with \textit{sctransform} or standard \textit{Seurat} normalization. \textbf{D:} Clustering accuracy across the normalization/dimensionality estimation methods.}
    \label{fig:figure8}
\end{figure}

\begin{figure}
    \centering
    \ifshowfigs
    \includegraphics[scale=0.43, keepaspectratio]{{main_figures/clustering.pdf}}
    \fi
    \caption{\textbf{Evaluation of clustering methods.} Clustering accuracy (\textbf{A}) and run time (\textbf{B}) of \textit{scran}- and \textit{Seurat}-based clustering tools in combination with \textit{sctransform} and standard \textit{Seurat}'s normalization.}
    \label{fig:figure9}
\end{figure}

\begin{figure}
    \centering
    \ifshowfigs
    \includegraphics[scale=0.55, keepaspectratio]{{main_figures/imputation.pdf}}
    \fi
    \caption{\textbf{Evaluation of imputation/denoising methods} Average silhouette width per subpopulation (\textbf{A}) and clustering accuracy (\textbf{B}) with or without (indicated as \textit{none}) application of a denoising/imputation method.}
    \label{fig:figure10}
\end{figure}

\begin{figure}[h]
    \centering
    \ifshowfigs
    \includegraphics[scale=0.9,keepaspectratio]{main_figures/pipeComp_summary.pdf}
    \fi
    \caption{\textbf{Recommendations of tools from filtering to clustering.}}
    \label{fig:figure11}
\end{figure}
